---
title: "Simulating Population Dynamics of Earthworms Using Markov Chains"
author: "Your Name"
output:
  pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Simulating Population Dynamics of Earthworms Using Markov Chains

## Introduction
Simulating population dynamics is crucial for understanding the adaptive capacity of earthworm populations under environmental pressures. This chapter employs a Markov chain approach to model the transitions between fitness categories and simulate fitness distributions over time. By incorporating environmental variables, the simulation captures the effects of habitat variations on earthworm populations.

## Markov Chain Model

### States and Transition Probabilities
Each earthworm population is categorized into fitness states, denoted as \( S_1, S_2, \dots, S_n \). The states represent discrete fitness levels, where higher states indicate greater fitness. Transitions between states are governed by a Markov chain with:
1. **State Transition Matrix (\(P\))**:
   \[
   P = \begin{bmatrix}
   p_{11} & p_{12} & \cdots & p_{1n} \\
   p_{21} & p_{22} & \cdots & p_{2n} \\
   \vdots & \vdots & \ddots & \vdots \\
   p_{n1} & p_{n2} & \cdots & p_{nn}
   \end{bmatrix},
   \]
   where \(p_{ij}\) is the probability of transitioning from state \(S_i\) to state \(S_j\).

2. **Fitness Transitions**:
   - Transitions depend on environmental factors (e.g., soil pH, moisture, temperature).
   - Transition probabilities are modified by a fitness function:
     \[
     p_{ij}(t) = p_{ij,0} \cdot F_{\text{env}}(t),
     \]
     where \(p_{ij,0}\) is the baseline transition probability, and \(F_{\text{env}}(t)\) reflects environmental fitness.

### Initial Population Distribution
The initial population is distributed across fitness states as:
\[
\mathbf{N}(0) = \begin{bmatrix} N_1(0) \\
N_2(0) \\
\vdots \\
N_n(0) \end{bmatrix},
\]
where \(N_i(0)\) is the number of individuals in state \(S_i\) at time \(t = 0\).

### Population Dynamics
The population distribution at time \(t\) evolves as:
\[
\mathbf{N}(t+1) = P \cdot \mathbf{N}(t).
\]

## Simulation Framework

### Defining Environmental Fitness Function
The environmental fitness function \(F_{\text{env}}(t)\) modifies transition probabilities:
\[
F_{\text{env}}(t) = F_{\text{pH}} \cdot F_{M} \cdot F_{T},
\]
where:
- \(F_{\text{pH}}\), \(F_{M}\), and \(F_{T}\) represent fitness contributions from pH, moisture, and temperature, respectively.

### Numerical Simulation
```{r simulation, echo=TRUE}
# Parameters
n_states <- 5  # Number of fitness states
n_steps <- 50  # Number of time steps
initial_population <- c(50, 30, 10, 5, 5)  # Initial distribution across states

# Baseline transition matrix
P <- matrix(c(
  0.6, 0.3, 0.1, 0.0, 0.0,
  0.2, 0.5, 0.2, 0.1, 0.0,
  0.1, 0.3, 0.4, 0.2, 0.0,
  0.0, 0.2, 0.3, 0.4, 0.1,
  0.0, 0.0, 0.1, 0.3, 0.6
), nrow = n_states, byrow = TRUE)

# Fitness modifiers (environmental effects)
environmental_fitness <- function(t) {
  # Example: Sinusoidal temperature effect
  temp_effect <- exp(-0.1 * (sin(2 * pi * t / n_steps) - 1)^2)
  moisture_effect <- exp(-0.05 * (runif(1, 0.8, 1.2) - 1)^2)
  pH_effect <- exp(-0.03 * (runif(1, 0.9, 1.1) - 1)^2)
  temp_effect * moisture_effect * pH_effect
}

# Simulate population dynamics
population <- matrix(0, nrow = n_steps, ncol = n_states)
population[1, ] <- initial_population

for (t in 2:n_steps) {
  fitness_mod <- environmental_fitness(t)
  modified_P <- P * fitness_mod
  modified_P <- sweep(modified_P, 1, rowSums(modified_P), FUN = "/")  # Normalize rows
  population[t, ] <- population[t - 1, ] %*% modified_P
}

# Visualization
matplot(1:n_steps, population, type = "l", lty = 1, col = 1:n_states,
        xlab = "Time Steps", ylab = "Population Size",
        main = "Fitness Distribution Over Time")
legend("topright", legend = paste("State", 1:n_states), col = 1:n_states, lty = 1)
```

## Applications and Implications

### Predicting Fitness Distributions
The model predicts how environmental changes affect the distribution of fitness categories over time, providing insights into population resilience.

### Adapting to Environmental Stressors
Simulations reveal which fitness states are most vulnerable to specific environmental pressures, guiding conservation strategies.

### Optimizing Habitat Management
By identifying optimal environmental conditions, the model aids in creating habitats that maximize population fitness.

## Conclusion
Using Markov chains to simulate earthworm population dynamics under environmental pressures provides a powerful tool for ecological research and conservation. By capturing transitions between fitness states and incorporating environmental factors, this approach offers valuable predictions for population management.
```

