---
title: "Adapting Environmental Variables for Earthworm Population Models"
author: "Your Name"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Adapting Environmental Variables for Earthworm Population Models

## Introduction
Environmental variables such as soil pH, moisture, and temperature significantly influence earthworm fitness, survival, and reproduction. This chapter integrates these spatial and temporal variables into population models, allowing for dynamic simulation of habitat variations and their impacts on earthworm populations.

## Defining Environmental Variables

### Key Environmental Factors
1. **Soil pH (\(\text{pH}\))**:
   - Affects enzyme activity, nutrient availability, and survival rates.
   - Modeled as a continuous variable influencing fitness \(F\):
     \[
     F_{\text{pH}} = \frac{1}{1 + e^{-k_1 (\text{pH} - \text{pH}_{\text{opt}})}},
     \]
     where \(k_1\) controls the sensitivity to deviations from optimal pH \(\text{pH}_{\text{opt}}\).

2. **Soil Moisture (\(M\))**:
   - Influences respiration and mobility.
   - Fitness dependency modeled as:
     \[
     F_{M} = e^{-k_2 (M - M_{\text{opt}})^2},
     \]
     where \(M_{\text{opt}}\) is the optimal moisture level, and \(k_2\) controls the penalty for deviations.

3. **Temperature (\(T\))**:
   - Governs metabolic rates and reproductive success.
   - Modeled using a Gaussian function:
     \[
     F_{T} = e^{-k_3 (T - T_{\text{opt}})^2},
     \]
     where \(T_{\text{opt}}\) is the optimal temperature, and \(k_3\) determines sensitivity.

### Combined Environmental Fitness Function
The combined fitness function incorporating all environmental factors is:
\[
F_{\text{env}} = F_{\text{pH}} \cdot F_{M} \cdot F_{T}.
\]
This product reflects the multiplicative influence of each variable on overall fitness.

## Dynamic Environmental Changes

### Temporal Dynamics
Environmental variables vary over time due to seasonal changes or anthropogenic factors. These dynamics are modeled as:
1. **Periodic Variations** (e.g., seasonal temperature changes):
   \[
   T(t) = T_{\text{avg}} + A_T \cdot \sin(\omega t + \phi),
   \]
   where:
   - \(T_{\text{avg}}\): Average temperature.
   - \(A_T\): Amplitude of seasonal fluctuation.
   - \(\omega\): Angular frequency.
   - \(\phi\): Phase offset.

2. **Stochastic Perturbations** (e.g., rainfall variability):
   \[
   M(t) = M_{\text{det}}(t) + \epsilon(t),
   \]
   where \(\epsilon(t)\) is a noise term drawn from a normal distribution \(N(0, \sigma^2)\).

### Spatial Variability
Spatial heterogeneity is introduced through environmental gradients:
\[
\text{pH}(x, y) = \text{pH}_{\text{avg}} + \nabla \text{pH} \cdot (x + y),
\]
where \(\nabla \text{pH}\) represents the gradient in soil pH across spatial coordinates \(x, y\).

## Modeling Population Dynamics Under Environmental Variation

### Modified Birth-Death Model
Environmental fitness \(F_{\text{env}}\) modifies survival and reproduction probabilities:
1. Survival probability for stage \(i\):
   \[
   p_i(t) = p_{i, 0} \cdot F_{\text{env}}(t),
   \]
   where \(p_{i, 0}\) is the baseline survival probability.

2. Reproduction rate:
   \[
   r_a(t) = r_{a, 0} \cdot F_{\text{env}}(t),
   \]
   where \(r_{a, 0}\) is the baseline reproduction rate.

### Simulation Framework
Simulate the effects of environmental variables on population dynamics:
```{r simulation, echo=TRUE}
# Parameters
n_steps <- 100
x_coords <- seq(0, 10, length.out = 100)
y_coords <- seq(0, 10, length.out = 100)
time <- 1:n_steps

# Environmental parameters
pH_opt <- 7
M_opt <- 0.3
T_opt <- 20
k1 <- 1
k2 <- 10
k3 <- 10

# Initialize fitness grids
pH <- outer(x_coords, y_coords, function(x, y) pH_opt + 0.1 * (x + y))
moisture <- M_opt + rnorm(length(x_coords) * length(y_coords), mean = 0, sd = 0.05)
temperature <- T_opt + 5 * sin(2 * pi * time / n_steps)

# Fitness calculation
fitness_env <- function(pH, M, T) {
  F_pH <- 1 / (1 + exp(-k1 * (pH - pH_opt)))
  F_M <- exp(-k2 * (M - M_opt)^2)
  F_T <- exp(-k3 * (T - T_opt)^2)
  F_pH * F_M * F_T
}

# Simulate fitness over time
fitness <- array(NA, dim = c(length(x_coords), length(y_coords), n_steps))
for (t in time) {
  fitness[,,t] <- fitness_env(pH, moisture, temperature[t])
}

# Visualize fitness at final time step
image(x_coords, y_coords, fitness[,,n_steps], main = "Fitness Landscape", xlab = "X", ylab = "Y", col = heat.colors(100))
```

## Applications and Implications

### Climate Change Scenarios
The model predicts how gradual shifts in temperature or moisture affect population dynamics, providing insights into potential range expansions or contractions.

### Soil Management Practices
By simulating spatial variability, the model identifies areas requiring intervention, such as pH adjustments or irrigation.

### Conservation Efforts
Temporal dynamics help forecast population resilience under fluctuating environmental conditions, aiding conservation planning.

## Conclusion
Incorporating environmental variables into population models enhances their ecological realism. This approach provides a powerful tool for predicting earthworm population responses to dynamic habitats, informing sustainable management and conservation efforts.
```

