---
title: "Implementation of Evolution Model for Earthworms"
author: "Adapted Framework"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
The study of earthworm population dynamics is critical for understanding their ecological roles, including soil fertility, nutrient cycling, and carbon sequestration. Evolutionary dynamics, particularly those involving mutation and preferential attachment, provide a framework for modeling population behaviors under environmental pressures. This document implements a stochastic birth-death model tailored to earthworms, integrating mathematical and ecological concepts for simulating their population dynamics.

## Background
Earthworm populations are influenced by ecological interactions, environmental stressors, and genetic variability. Mutation introduces variability in traits such as reproduction and stress resistance, while preferential attachment ensures that individuals with higher fitness contribute more significantly to the population.

### Key Concepts
1. **Mutation:** Mutations introduce variability by altering fitness values, such as reproduction rate or stress resistance. Mathematically, mutation can be expressed as:
   \[
   F_i = F_0 + \Delta F,
   \]
   where \(F_0\) is the baseline fitness, and \(\Delta F\) represents the fitness change due to mutation.

2. **Birth and Preferential Attachment:** Birth events occur with probability \(p\). In the case of non-mutant births, fitness \(F_j\) is chosen proportional to the population size \(N_j\) at fitness level \(j\):
   \[
   P(F_j) \propto N_j.
   \]

3. **Death:** Death events occur with probability \(1 - p\), removing individuals from the least fit group to simulate "survival of the fittest."

### Related Work
The model implemented in this document builds on ecological and evolutionary principles:
- **Edwards and Bohlen (1996):** Explored the biology and ecology of earthworms, providing foundational insights into their population dynamics.
- **Lavelle and Spain (2001):** Highlighted the ecological roles of earthworms in nutrient cycling and soil structure.
- **Curry and Schmidt (2007):** Reviewed earthworm responses to environmental stressors such as heavy metals and pesticides.

## Objectives
The primary goal of this implementation is to simulate the evolutionary dynamics of earthworm populations, incorporating:
- Mutation-driven variability.
- Preferential attachment mechanisms.
- Birth-death stochastic processes.

This model provides a quantitative framework for analyzing population behaviors under varying ecological and genetic scenarios, aiding in soil health assessments and conservation efforts.

### References
- Edwards, C. A., & Bohlen, P. J. (1996). *Biology and Ecology of Earthworms.* Springer.
- Lavelle, P., & Spain, A. V. (2001). *Soil Ecology.* Kluwer Academic Publishers.
- Curry, J. P., & Schmidt, O. (2007). The feeding ecology of earthworms – A review. *Pedobiologia, 50*(6), 463–477.

# Parameters
Set the key parameters for the simulation.

```{r parameters}
# Load libraries
library(ggplot2)
library(dplyr)
library(tidyr)

# Define parameters
set.seed(123)
p <- 0.6  # Probability of birth
r <- 0.4  # Probability of mutation
n_steps <- 100  # Number of time steps
initial_population <- data.frame(fitness = c(0.3, 0.6, 0.9), size = c(10, 20, 5))
```

# Simulation Framework
The simulation follows a discrete-time process with the following rules:

1. **Birth**: With probability `p`, an individual is born.
   - With probability `r`, the individual is a mutant with a random fitness value.
   - Otherwise, preferential attachment determines its fitness.

2. **Death**: With probability `1 - p`, the least fit individual dies.

## Simulation Function
Define a function to simulate the population dynamics over time.

```{r simulation-function}
simulate_population <- function(initial_population, n_steps, p, r) {
  population <- initial_population
  results <- list()

  for (t in 1:n_steps) {
    if (runif(1) < p) {
      # Birth event
      if (runif(1) < r) {
        # Mutation: New fitness level uniformly in [0, 1]
        new_fitness <- runif(1)
        population <- population %>%
          add_row(fitness = new_fitness, size = 1)
      } else {
        # Preferential attachment
        total_size <- sum(population$size)
        probabilities <- population$size / total_size
        chosen <- sample(seq_len(nrow(population)), size = 1, prob = probabilities)
        population$size[chosen] <- population$size[chosen] + 1
      }
    } else {
      # Death event: Remove one individual from the least fit group
      least_fit <- which.min(population$fitness)
      if (population$size[least_fit] > 1) {
        population$size[least_fit] <- population$size[least_fit] - 1
      } else {
        population <- population[-least_fit, ]
      }
    }

    # Store the population state
    results[[t]] <- population
  }

  results
}
```

# Running the Simulation
Simulate the evolution of the population using the defined parameters and function.

```{r run-simulation}
# Run simulation
simulation_results <- simulate_population(initial_population, n_steps, p, r)

# Prepare results for visualization
plot_data <- bind_rows(lapply(seq_along(simulation_results), function(t) {
  data <- simulation_results[[t]]
  data$time <- t
  data
}))
```

# Visualization
## Fitness Distribution Over Time
Visualize how the fitness distribution evolves across the simulation timeline.

```{r plot-fitness-distribution}
ggplot(plot_data, aes(x = fitness, y = size, color = as.factor(time))) +
  geom_line() +
  labs(title = "Fitness Distribution Over Time",
       x = "Fitness",
       y = "Population Size",
       color = "Time Step") +
  theme_minimal()
```

## Population Dynamics
### Total Population
Plot the total population size over time.

```{r plot-total-population}
summary_data <- plot_data %>%
  group_by(time) %>%
  summarise(total_population = sum(size),
            mean_fitness = weighted.mean(fitness, size))

ggplot(summary_data, aes(x = time, y = total_population)) +
  geom_line() +
  labs(title = "Total Population Over Time",
       x = "Time Step",
       y = "Total Population") +
  theme_minimal()
```

### Mean Fitness
Plot the mean fitness of the population over time.

```{r plot-mean-fitness}
ggplot(summary_data, aes(x = time, y = mean_fitness)) +
  geom_line() +
  labs(title = "Mean Fitness Over Time",
       x = "Time Step",
       y = "Mean Fitness") +
  theme_minimal()
```

# Conclusion
This implementation provides a framework for simulating and visualizing the evolution of earthworm populations. The integration of mutation and preferential attachment mechanisms highlights key ecological dynamics, making this model relevant for soil health studies and conservation efforts.
